06_gam

### 2024-01-24 06_gam.Rmd
Update 06_gam.Rmd
added distance to escarpment to GAMs

Load allDrifts object, split out Cuvier's and Sperm whales
```{r}
load(file = "data/allDrifts.rda")

ZC_filt <- allDrifts %>% 
  filter(species %in% c("ZC", NA))

PM_filt <- allDrifts %>% 
  filter(species %in% c("PM", NA))

@@

```{r}
# Environ only - binomial
ZC_GAM1 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
              data = ZC_filt, family = binomial, method = "REML")
summary(ZC_GAM1)
gam.check(ZC_GAM1)

# Environ + noise - binomial
ZC_GAM2<- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
                      data = ZC_filt, family = binomial, method = "REML")
summary(ZC_GAM2)
gam.check(ZC_GAM2)

# Environ only - poisson
ZC_GAM3 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
              data = ZC_filt, family = poisson, method = "REML")
summary(ZC_GAM3)
gam.check(ZC_GAM3)

# Environ + noise - poisson
ZC_GAM4<- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
                      data = ZC_filt, family = poisson, method = "REML")
summary(ZC_GAM4)
gam.check(ZC_GAM4)
plot(ZC_GAM4, all.terms = TRUE, scale = 0)

# Environ only - gaussian
ZC_GAM5 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
                  data = ZC_filt, method = "REML")
summary(ZC_GAM5)
gam.check(ZC_GAM5)

# Environ + noise - gaussian
ZC_GAM6 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
                  data = ZC_filt, method = "REML")
summary(ZC_GAM6)
gam.check(ZC_GAM6)

# Environ only - Tweedie
ZC_GAM7 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
                  data = ZC_filt, family = tw(link = "log"), method = "REML")
summary(ZC_GAM7)
gam.check(ZC_GAM7)

# Environ + noise - Tweedie
ZC_GAM8 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
                  data = ZC_filt, family = tw(link = "log"), method = "REML")
summary(ZC_GAM8)
gam.check(ZC_GAM8)

plot(ZC_GAM1, all.terms = TRUE, scale = 0, pages = 1)
plot(ZC_GAM2, all.terms = TRUE, scale = 0, pages = 2)
plot(ZC_GAM3, all.terms = TRUE, scale = 0, pages = 1)
plot(ZC_GAM4, all.terms = TRUE, scale = 0, pages = 2)
plot(ZC_GAM5, all.terms = TRUE, scale = 0, pages = 1)
plot(ZC_GAM6, all.terms = TRUE, scale = 0, pages = 2)
plot(ZC_GAM7, all.terms = TRUE, scale = 0, pages = 1)
plot(ZC_GAM8, all.terms = TRUE, scale = 0, pages = 2)
AIC(ZC_GAM4, ZC_GAM8)
```

@@

```{r}
# Environ only - binomial
PM_GAM1 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + 
                    s(sst_mean),
              data = PM_filt, family = binomial, method = "REML")
summary(PM_GAM1)
gam.check(PM_GAM1)

# Environ + noise - binomial
PM_GAM2 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + 
                 s(sst_mean) + s(TOL_125) + s(TOL_2000),
              data = PM_filt, family = binomial, method = "REML")
summary(PM_GAM2)
gam.check(PM_GAM2)

# Environ only - poisson
PM_GAM3 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean),
              data = PM_filt, family = poisson, method = "REML")
summary(PM_GAM3)
gam.check(PM_GAM3)

# Environ + noise - poisson
PM_GAM4 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) 
               + s(sst_mean) + s(TOL_125) + s(TOL_2000),
              data = PM_filt, family = poisson, method = "REML")
summary(PM_GAM4)
gam.check(PM_GAM4)

# Environ only - gaussian
PM_GAM5 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean),
              data = PM_filt, method = "REML")
summary(PM_GAM5)
gam.check(PM_GAM5)

# Environ + noise - gaussian
PM_GAM6 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean)
               + s(TOL_125) + s(TOL_2000),
              data = PM_filt, method = "REML")
summary(PM_GAM6)
gam.check(PM_GAM6)

# Environ only - Tweedie
PM_GAM7 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean),
              data = PM_filt, family = tw(link = "log"), method = "REML")
summary(PM_GAM7)
gam.check(PM_GAM7)

# Environ + noise - Tweedie
PM_GAM8 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean)
               + s(TOL_125) + s(TOL_2000),
              data = PM_filt, family = tw(link = "log"), method = "REML")
summary(PM_GAM8)
gam.check(PM_GAM8)

plot(PM_GAM1, all.terms = TRUE, scale = 0)
plot(PM_GAM2, all.terms = TRUE, scale = 0)
plot(PM_GAM3, all.terms = TRUE, scale = 0)
plot(PM_GAM4, all.terms = TRUE, scale = 0)
plot(PM_GAM5, all.terms = TRUE, scale = 0)
plot(PM_GAM6, all.terms = TRUE, scale = 0)
plot(PM_GAM7, all.terms = TRUE, scale = 0)
plot(PM_GAM8, all.terms = TRUE, scale = 0)
AIC(PM_GAM4, PM_GAM8)

@@

PM_filt2 <- drift_escarp %>% 
  filter(species %in% c("PM", NA))
#####

# Poisson
ZC_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
              data = ZC_filt2, family = poisson, method = "REML")
summary(ZC_GAM_escarp)
plot(ZC_GAM_escarp, all.terms = TRUE)

# Tweedie (family = tw(link = "log"))
ZC_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
              data = ZC_filt2, family = tw(link = "log"), method = "REML")
summary(ZC_GAM_escarp)
plot(ZC_GAM_escarp, all.terms = TRUE)

# Poisson
PM_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(dist2slope) + s(mldDepth) + s(sst_mean)
               + s(TOL_125) + s(TOL_2000),
              data = PM_filt2, family = poisson, method = "REML")
summary(PM_GAM_escarp)
plot(PM_GAM_escarp, all.terms = TRUE)
plot.gam(PM_GAM_escarp)

# Tweedie
PM_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(dist2slope) + s(mldDepth) + s(sst_mean)
               + s(TOL_125) + s(TOL_2000),
              data = PM_filt2, family = tw(link = "log"), method = "REML")
summary(PM_GAM_escarp)
```

>>>>

Load drift_escarp object, split out Cuvier's and Sperm whales
```{r}
load(file = "D:/Buoy_dfs/drift_escarpment.rda")

drift_escarp <- drift_escarp %>%
  st_drop_geometry()

ZC_filt <- drift_escarp %>% 
  filter(species %in% c("ZC", NA))

PM_filt <- drift_escarp %>% 
  filter(species %in% c("PM", NA))

@@

```{r}
# Environ only - binomial
ZC_GAM1 <- gam(Wpresence>0 ~ s(depth) + s(ttTemp) + s(ssh_mean),
              data = ZC_filt, family = binomial, 
              method = "REML")
summary(ZC_GAM1)
gam.check(ZC_GAM1)

# Environ + noise - binomial
ZC_GAM2<- gam(Wpresence>0 ~ s(depth) + s(ttTemp) + s(ssh_mean) + 
                s(TOL_125) + s(TOL_2000),
                      data = ZC_filt, family = binomial, 
              method = "REML")
summary(ZC_GAM2)
gam.check(ZC_GAM2)

plot(ZC_GAM1, all.terms = TRUE, scale = 0, pages = 1)
plot(ZC_GAM2, all.terms = TRUE, scale = 0, pages = 3)

# Compare AIC scores
AIC(ZC_GAM1, ZC_GAM2)
```

@@

```{r}
# Environ only - binomial
PM_GAM1 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + 
                    s(dist2escarp),
              data = PM_filt, family = binomial, method = "REML")
summary(PM_GAM1)
gam.check(PM_GAM1)

# Environ + noise - binomial
PM_GAM2 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + 
                 s(dist2escarp) + s(TOL_125) + s(TOL_2000),
              data = PM_filt, family = binomial, method = "REML")
summary(PM_GAM2)
gam.check(PM_GAM2)

# Plot
plot(PM_GAM1, all.terms = TRUE, scale = 0, pages = 1)
plot(PM_GAM2, all.terms = TRUE, scale = 0, pages = 1)

# compare AIC scores
AIC(PM_GAM1, PM_GAM2)

@@

PM_filt2 <- drift_escarp %>% 
  filter(species %in% c("PM", NA))
#####
```
